---
title: "R Analysis"
author: "Jennifer Chang"
date: "1/8/2021"
output:
  github_document
editor_options:
  chunck_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "imgs/2_DE_analysis"
)
```

## R Markdown

This is a placeholder to store any analysis in R (edgeR, DESeq2). Also a way to look at the underlying assumptions (RNASeq distribution functions and how it behaves in an analysis pipeline. Is the distribution realistic?)

## Install any required libraries

```{r, warning=FALSE, message=FALSE}
# == Install any CRAN packages
# cran_pkgs <- c("QuasiSeq")
# install.packages(cran_pkgs)

# == Install any Bioconductor packages
# bioc_pkgs <- c("edgeR", "baySeq", "DESeq2", "ballgown", "NOISeq", "limma")
# BiocManager::install(bioc_pkgs)

# ====================   Year of development
library(tidyverse)
library(magrittr)
library(edgeR)         # 2010
library(baySeq)        # 2010
library(QuasiSeq)      # 2012
library(DESeq2)        # 2014    # <= actually a bit strange that DESeq2 & edgeR tend to be equivalent, despite being 4yrs later
library(ballgown)      # 2014
library(NOISeq)        # 2015
library(limma)         # 2015
```

<!--  ## For now comment out exploration portion.
## Distribution

RNASeq is assumed to be [negatively binomially distributed](https://en.wikipedia.org/wiki/Negative_binomial_distribution) according to the edgeR and DESeq2 papers. 

$$
Y_{gi} \sim NB(M_{i}p{gj}, \phi_{g})
$$

What does a negative binomial distribution look like?

```{r negbinomial, fig.width=6, fig.height=4}
y  <- rnbinom( n = 1000,       # Randomly generate 1000 entries (number of replicates?)
               size = 5,       # at least 5 successes (expressed/not?)
               prob = 0.5      # probability of success in each case...maybe length of genome may affect this
               )

df <- data.frame(x = c(1:1000),
                 y = y)
                  
ggplot(data = df, aes(x=y)) + 
  geom_histogram(binwidth = 1) +
  theme_bw() +
  labs(title="Negative Binomial Distribution (n=1000, size=5, prob=0.5)") +
  geom_vline(xintercept=5, color="blue")
```

Notice how the peak of the distribution is around 5 (`size=5`). Play around with the `rnbinom` parameters, how does that change the shape of the distribution? What implications does this have for RNASeq to determine if a gene is differentially expressed across two treatments?
-->

# Load in RNAseq counts

Load in the RNAseq counts generated by GSNAP.

```{r}
data_maize <- read_delim("maize_genecounts.txt", delim = "\t")
data_bee <- read_delim("bee_genecounts.txt", delim="\t")
```

# Diagnostic plots

These are quick plots to look at the data and notice if any of the counts seem unusally high or low, or multimodal. It can indicate a problem with a sample, but does not necessarily mean there is a problem with a sample.

## Bee

```{r bee_linechart, fig.width=8, fig.height=5}
melt_bee <- data_bee %>%
  pivot_longer(!Geneid, names_to = "replicate", values_to = "count") %>%
  mutate(
    group=substr(replicate, 1, 3)                 # <= replace this with metadata information later
  )

melt_bee %>% 
  ggplot(., aes(x=replicate, y=count, group=Geneid)) + 
  geom_line(alpha=0.5) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle=90, vjust=0.5)
  ) +
  labs(title="GSNAP RNAseq Gene Profiles")
```

Distribution, violin plot

```{r violin, fig.width=11, fig.height=8}
melt_bee %>% 
  ggplot(., aes(x=replicate,y=log(count))) +       # <= log transformed as a rough normalization... we will use DESeq2's normalization eventually
  geom_violin(alpha=0.5) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle=90)
  ) +
  labs(title="GSNAP RNAseq Gene Profiles") +
  facet_wrap(~group, scales="free", drop=T)
```

Check if any of the replicates seem obviously different from the others. Mostly seems fine, `1-E07-F3-S59 distribution` is a little low compared to others, `1-B03-A16_S21` seems a little high compared to others.

## Maize

```{r maize_linechart, fig.width=8, fig.height=5}
melt_maize <- data_maize %>%
  pivot_longer(!Geneid, names_to = "replicate", values_to = "count")

melt_maize %>% 
  ggplot(., aes(x=replicate, y=count, group=Geneid)) + 
  geom_line(alpha=0.5) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle=90, vjust=0.5)
  ) +
  labs(title="GSNAP RNAseq Gene Profiles")
```

```{r maize_violin, fig.width=8, fig.height=4}
melt_maize %>% 
  ggplot(., aes(x=replicate,y=log(count))) +       # <= log transformed as a rough normalization... we will use DESeq2's normalization eventually
  geom_violin(alpha=0.5) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle=90)
  ) +
  labs(title="GSNAP RNAseq Gene Profiles")
#  facet_wrap(~group, scales="free", drop=T)
```
Check if any of the replicates seem obviously different from the others. Mostly seems okay, `SRR1573509` and `SRR1573523` is a little low compared to others, `SRR1573513` seems a little high compared to others.